\documentclass[11pt,letterpaper]{article}
\usepackage{fullpage}
\usepackage[pdftex]{graphicx}
\usepackage{amsfonts,eucal,amsbsy,amsopn,amsmath}
\usepackage{url}
\usepackage[sort&compress]{natbib}
\usepackage{natbibspacing}
\usepackage{latexsym}
\usepackage{wasysym} 
\usepackage{rotating}
\usepackage{fancyhdr}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\usepackage[dvipsnames,usenames]{color}
\usepackage{multicol}
\definecolor{orange}{rgb}{1,0.5,0}
\usepackage{caption}
\renewcommand{\captionfont}{\small}
\setlength{\oddsidemargin}{-0.04cm}
\setlength{\textwidth}{16.59cm}
\setlength{\topmargin}{-0.04cm}
\setlength{\headheight}{0in}
\setlength{\headsep}{0in}
\setlength{\textheight}{22.94cm}
\newcommand{\ignore}[1]{}
\newenvironment{enumeratesquish}{\begin{list}{\addtocounter{enumi}{1}\arabic{enumi}.}{\setlength{\itemsep}{-0.25em}\setlength{\leftmargin}{1em}\addtolength{\leftmargin}{\labelsep}}}{\end{list}}
\newenvironment{itemizesquish}{\begin{list}{\setcounter{enumi}{0}\labelitemi}{\setlength{\itemsep}{-0.25em}\setlength{\labelwidth}{0.5em}\setlength{\leftmargin}{\labelwidth}\addtolength{\leftmargin}{\labelsep}}}{\end{list}}

\bibpunct{(}{)}{;}{a}{,}{,}
\newcommand{\nascomment}[1]{\textcolor{blue}{\textbf{[#1 --NAS]}}}


\pagestyle{fancy}
\lhead{}
\chead{}
\rhead{}
\lfoot{}
\cfoot{\thepage~of \pageref{lastpage}}
\rfoot{}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}


\title{11-712:  NLP Lab Report\\Slovene}
\author{Matt Gardner}
\date{April 26, 2013}

\begin{document}
\maketitle
\begin{abstract}
\nascomment{one paragraph here summarizing what the paper is about}
\end{abstract}

\nascomment{brief introduction}

\section{Basic Information about Slovene}

Slovene is a Slavic language spoken in the country of Slovenia.  It is spoken
natively by around 2 million people, most of whom live in Slovenia.  Slovene is
one of the official languages of the European Union, so there is a moderate
amount of data available for computational linguists.  Slovene's closest
relative is Serbo-Croatian, being mutually intelligible with some variants of
Croatian.  As with many Slavic languages, Slovene is highly inflected and has
relatively free word order~\citep{wikipedia-slovene}.

\section{Past Work on the Morphology of Slovene}

There has been some work on the Slovene language by linguists.  There is a
grammar written by \cite{greenberg-2006-slovene-grammar} that gives some 75
pages to the morphology of the language.  There is also a more comprehensive
grammar written by \cite{herrity-2000-slovene-grammar}.  There are a few other
papers on Slovene morphology to be found in English by searching Google scholar
(e.g., \cite{bidwell-1969-outline-slovene-morphology}), and more that is
written in Slovene.

As far as work in computational linguists on Slovene, a small body of
literature exists.  Toma\v{z} Erjavec, from the Jo\v{z}ef Stefan Institute,
published a description of an early morphological analyzer for Slovene.  The
paper is only a two-and-a-half page, very high-level description, and the
system was implemented in Prolog on a VAX
machine~\citep{erjavec-1990-slovene-analyzer}.  Erjavec also was a main
contributor to the Slovene Dependency Treebank, which includes morphological
information and was a part of the CoNLL-X Shared Task on multilingual
dependency parsing in 2006~\citep{dvzeroski-2006-sdt}.

More recently, there was a Slovene morphological analyzer released on
SourceFourge in May of 2012.  This analyzer uses a maximum entropy model and
achieves 92.5\% accuracy~\citep{grcar-2012-obeliks}.  The system (called
Obeliks) was trained on a corpus of five hundred thousand words.

\section{Available Resources}

The available resources for work on the Slovene language consist largely of
things mentioned in the previous section.  I already listed reference grammars,
the Slovene dependency treebank, and a corpus available with the Obeliks tool.
In addition, there is a Slovene lexicon available with the MULTEXT-East
dataset~\citep{erjavec-1998-slovene-lexicon}; I have requested access to the
lexicon and hope to have it soon.  Assuming I get access to the lexicon, I will
use the lexicon as I develop the analyzer, and I will use the morphology labels
from the Slovene Dependency Treebank and from the dataset released with the
Obeliks system as my two test sets.  These two datasets have hundreds of
thousands of tokens (though I have not yet counted the number of word types),
and the lexicon has over 15,000 lemmas, with full inflectional paradigms for
each lemma.

\section{Survey of Phenomena in Slovene}

For the phenomena my morphological analyzer will cover, I will follow exactly
the morphosyntactic specifications given by~\cite{erjavec-mds}.  This
specification gives twelve parts of speech.  Each part of speech has a set of
features or attributes that are marked in words of that part of speech.  These
are the parts of speech, with their features:

\begin{itemize}
  \item \textbf{Noun}: type (common / proper), gender
    (masculine / feminine / neuter), number (singular / dual / plural),
    case (nominative / genitive / dative / accusative / locative /
    instrumental), animate (yes / no)
  \item \textbf{Verb}: type (main / auxiliary), aspect (perfective /
    progressive / biaspectual), form (infinitive / supine / participle /
    present / future / conditional / imperative), person (first / second /
    third), number, gender, negative (yes / no)
  \item \textbf{Preposition}: case (specifies the case it governs, not the case
    it is inflected for)
  \item \textbf{Conjunction}: type (coordinating / subordinating)
  \item \textbf{Adjective}: type (general / possessive / participle), degree
    (positive / comparative / superlative), gender, number, case, definiteness
    (yes / no)
  \item \textbf{Pronoun}: type (personal / possessive / demonstrative /
    relative / reflexive / general / interrogative / indefinite / negative),
    person, gender, number, case, owner number (for possessive pronouns), owner
    gender, clitic (yes / bound)
  \item \textbf{Adverb}: type (general / participle), degree
  \item \textbf{Numeral}: form (digit / roman / letter), type (cardinal /
    ordinal / pronominal / special), gedner, number, case, definiteness
  \item \textbf{Particle}: (no features)
  \item \textbf{Interjection}: (no features)
  \item \textbf{Abbreviation}: (no features)
  \item \textbf{Residual}: type (foreign / typo / program)
\end{itemize}

This list is roughly in the order I would assign them priority.  I doubt that I
would be able to handle open class abbreviations or interjections, or the
loosely-defined ``residual'' part of speech.  The closed classes should be
relatively easy, as I have a large corpus that has been labeled with these
morphosyntactic descriptions, so I will only need to produce a few small
lexicons for these words.  I will probably not try to handle common nouns vs.
proper nouns at first, nor to distinguish main verb vs. auxiliary verb (which
depends on context).  Aspect of verbs will also be difficult, and so I will
save it for later, and there are a few features listed above that only rarely
show up in word forms (like noun animacy and verb negativity) which could be
ignored at the beginning.

\section{Initial Design}

The morphological analyzer is designed as follows.  I am using as my working
data set a large lexicon generously provided by Toma\v{z} Erjavec.  This
lexicon has inflected word forms together with their lemma and morphosyntactic
disambiguation, so it is a perfect test set for my morphological analyzer.  I
processed the lexicon and split it into 12 individula lexicons, one for each
main part of speech listed in the previous section, using the lemma provided as
the lexical entry.  I then created 12 individual test sets, again one for each
part of speech, grouping all analyses of an individual form.  I wrote automated
tools to test each part of speech independently, using the lexicon and test set
for that part of speech, together with the inflection rules written for each
part of speech in separate files.  The organization of the system, then, is the
following:

\begin{itemize}
  \item The base lexicon and automatic processing files:
    \begin{itemize}
      \item The base lexicon, with approximately 3 million possible analyses of
        approximately 1 million word forms.
      \item A python script that processes the base lexicon and produces 12
        individual lexicons, one for each part of speech.  This script does
        some simple analysis of the lemmas to determine proper continuation
        lexicons for each word, when that is necessary (e.g., adjectives that
        end in -ev, -o\v{c}/-e\v{c}, and -i each have their own contiuation
        classes, because they have different analyses).
      \item Anther python script that processes the base lexicon and produces
        12 test files, one for each part of speech, and a single test file for
        all parts of speech at the same time (this separation is helpful, for
        instance, when a single word could have more than one part of speech; I
        can see how my analyzer works on adjectives without having confounding
        results from forms that could also be nouns).
    \end{itemize}
  \item The hand-written lexicon files:
    \begin{itemize}
      \item A base file that gives all of the possible parts of speech, and
        multicharacter symbols.
      \item For each part of speech, a rules file that adds endings and the
        inflectional meaning of those endings.
    \end{itemize}
  \item The FOMA file that specifies more complicated replacement rules, for
    stem changes and other phenomena.
\end{itemize}

This initial system design is able to correctly analyze between 75 and 95\% of
the word forms of the open class nouns, adjectives and verbs, and all of the
closed class prepositions.  Continued development can easily proceed in a
modular fashion, either by improving the lexicon for an individual part of
speech (both in the script that determines continuation classes for each lemma,
and in the rule lexicon for that part of speech), or by improving the stem
changing rules in the FOMA file.

\section{System Analysis on Corpus A}

\section{Lessons Learned and Revised Design}

\section{System Analysis on Corpus B}

\section{Final Revisions}

\section{Future Work}





\bibliographystyle{plainnat}
\bibliography{bib}
\label{lastpage}
\end{document}
